\Title

### 2021-02-22 latex/paper/main.tex
Edits to the VNHC section

\subsection{Notation}
We use the following notation in this paper. 
For \(T > 0\), the set of real numbers modulo is denote \(\Rt{T}\), with
\(\Rt{\infty} := \R\).
The gradient of the matrix-valued function 
\(A : \R^m \rightarrow \R^{n\times n}\) is the matrix of stacked partial
derivatives, 
\(\nabla_xA := (\pdiff{A}{x_1},\ldots,\pdiff{A}{x_m})\tpose \in \R^{nm \times n}\).
Given two matrices \(A \in \R^{n \times m}\) and \(B \in \R^{r \times s}\), the
Kronecker product \cite{kronprod} \(A \otimes B \in \R^{nr \times ms}\) is the
matrix
\[
    A \otimes B = \begin{bmatrix}
        A_{1,1}B & \cdots & A_{1,m} B \\
        \vdots & \ddots & \vdots \\
        A_{n,1} B & \cdots & A_{n,m} B
    \end{bmatrix} 
    .
\]

@@

\subsection{Simply Actuated Hamiltonian Systems}
Take an underactuated mechanical system modelled with generalized coordinates 
\(q = (q_1, \ldots, q_n)\) on the configuration manifold
\(\mathcal{Q} = \Rt{T_1} \times \cdot \Rt{T_n}\), where
\(T_i = 2\pi\) if \(q_i\) is an angle and \(T_i = \infty\) if \(q_i\) is a
length. The corresponding velocities are 
\(\dot{q} = (\dot{q}_1,\ldots,\dot{q}_n) \in \R^n\).

Suppose this system has Lagrangian
\(\mathcal{L}(q,\dot{q}) = 1/2 \dot{q}^T D(q) \dot{q} - P(q)\),
where \(P : \mathcal{Q} \rightarrow \mathbb{R}\) is the potential energy, and
the inertia matrix \(D : \mathcal{Q} \rightarrow \mathbb{R}^{n \times n}\)
is positive definite for all \(q \in \mathcal{Q}\).
By defining the \textit{conjugate of momentum} 
\(p_i := \pdiff{\mathcal{L}}{\dot{q}_i}\) so that the vector 
\(p = (p_1,\ldots,p_n) \in \R^n\) is given by \(p = D(q)\dot{q}\),
we find the \textit{Hamiltonian} of the system in \((q,p)\) coordinates
is\cite{landau_mechanics}
\begin{equation}\label{eqn:hamiltonian}
    \mathcal{H}(q,p) = 1/2 p\tpose D(q) p + P(q).
\end{equation}
The dynamics of the system in Hamiltonian form are
\begin{equation}\label{eqn:hamiltonian-eom-general}
    \begin{cases}
        \dot{q} = \nabla_p\mathcal{H} 
        , \\
        \dot{p} = -\nabla_q\mathcal{H} + B(q) \tau
        ,
    \end{cases}
\end{equation}
where \(\tau \in \R^k\) is a vector of generalized input forces and the input
matrix \(B : \mathcal{Q} \rightarrow \R^{n \times k}\) is full rank for all 
\(q \in \mathcal{Q}\).

@@

Because \(\tau\) is transformed by \(B(q)\), it is not in general clear how any
particular input force \(\tau_i\) will affect the dynamics of the system. 
The following definition provides a class of systems where the effect of input
forces is made explicit.

\begin{defn}
    Let \(\mathcal{H}\) be an \(n\)-DOF Hamiltonian system 
    with \(k \leq n\) actuators. 
    A set of canonical coordinates \((q,p)\) for this system
    are said to be \textit{simply actuated coordinates} if the
    input matrix \(B(q) \in \R^{n \times k}\) is of the form
    \[
        B(q) = \simpleB    
        .
    \]

\end{defn}

The first \((n-k)\) coordinates, labelled \(q_u\), are called the
\textit{unactuated coordinates}. The remaining \(k\) coordinates, labelled
\(q_a\), are called the \textit{actuated coordinates}. When grouping them
together, we will always put them in the order \((q_u, q_a)\) to fit with 
the definition. 
The corresponding \((p_u, p_a)\) are called the \textit{unactuated} and 
\textit{actuated momenta}, respectively.

Under the following assumptions on the input matrix, we will show that there is
a canonical transformation of ~\eqref{eqn:hamiltonian} into simply actuated
coordinates.

@@

In light of Lemma \ref{lemma:B-orthogonal}, there is no loss of generality
by making the following assumption.
\begin{assm}\label{assm:B-orthogonal}
    Assume that the input matrix \(B\) is
    left semi-orthogonal, i.e.~\(B\tpose B = I_k\). 
\end{assm}

Let now \(\mathbf{B} \in \R^{n\times n}\) be the following matrix:
\[
    \mathbf{B} = 
    \begin{bmatrix}
        B^\perp \\
        B\tpose \\
    \end{bmatrix}
    .
\]
Since \(B^\perp\) is a left annihilator of \(B\) and both \(B^\perp\) and
\(B\tpose\) are right semi-orthogonal, it is easy to show that \(\mathbf{B}\) is
orthogonal.
\begin{proof}
\[
    \mathbf{B}\mathbf{B}\tpose = 
    \begin{bmatrix}
        B^\perp (B^\perp)\tpose & B^\perp B \\
        (B^\perp B)\tpose & B\tpose B
    \end{bmatrix} = \Id{n}
    .
\]
Hence, \(\mathbf{B}\) is invertible with \(\mathbf{B}\inv = \mathbf{B}\tpose\).
\end{proof}

The following theorem shows that \(\mathbf{B}\) provides a canonical
transformation into simply actuated coordinates, so that only the actuated momenta
are affected by the input forces.

\begin{thm}\label{thm:simply-actuated}
    Under Assumptions \ref{assm:B-const},\ref{assm:B-perp}, and
    \ref{assm:B-orthogonal}, the Hamiltonian system ~\eqref{eqn:hamiltonian}
    has simply actuated canonical coordinates 
    \(\left(\tilde{q} = \mathbf{B}q, \tilde{p} = \mathbf{B}p\right)\). 
    The resulting dynamics are given by 
    \begin{gather}\label{eqn:simple-hamiltonian}
        \mathcal{H}(\tilde{q},\tilde{p}) = 
        \frac{1}{2} \tilde{p}\tpose \Minv(\tilde{q}) \tilde{p} + V(\tilde{q})
        , \\
       \begin{cases}
           \dot{\tilde{q}} = \Minv(\tilde{q})\tilde{p}
           , \\
           \dot{\tilde{p}} = -\frac{1}{2} (\Id{n} \otimes \tilde{p}\tpose)
           \nabla_{\tilde{q}} \Minv(\tilde{q}) \tilde{p}
           - \nabla_{\tilde{q}} V(\tilde{q}) + \simpleB \tau
            ,
        \end{cases} \nonumber
    \end{gather}
    where 
    \(\Minv(\tilde{q}) := \mathbf{B}D^{-1}(\mathbf{B}\tpose
    \tilde{q})\mathbf{B}\tpose\)
    and
    \(V(\tilde{q}) := P(\mathbf{B}\tpose \tilde{q})\).
\end{thm}
\begin{proof}
    \textbf{TODO: prove this in a terse way.}
\end{proof}
Throughout the rest of this paper, we will assume the system we are studying is
in simply actuated coordinates.

@@

\begin{defn}
    A \textit{virtual nonholonomic constraint} (VNHC) \textit{of order \(k\)} is a
    relation \(h(q,p) = 0\) where \(h : \mathcal{Q}\times\R^n \rightarrow \R^k\) is
    \(C^2\), \(\rank{\left[ dh_q,\, dh_p \right]} = k\) for all 
    \((q,p) \in h\inv(0)\), and there exists a feedback controller \(\tau(q,p)\)
    rendering the \textit{constraint manifold} \(\Gamma\) invariant,
    where
    \[
        \Gamma = \left\{(q,p) \mid h(q,p) = 0, dh_q \dot{q} + dh_p \dot{p} = 0\right\}
        .
    \]
\end{defn}

From the definition of VNHCs, one finds that the constraint manifold \(\Gamma\)
is a \(2(n-k)\)-dimensional closed embedded submanifold of 
\(\mathcal{Q} \times \R^n\).

%TODO: tie this together

\begin{thm}\label{thm:vnhc-regularity}
    A relation \(h(q,p) = 0\) for system ~\eqref{eqn:simply-actuated-hamiltonian}
    is a regular VNHC of order \(k\) if and only if \(dh_{p_a} = 0\) 
    and
    \[
        \rank{\left(dh_q \Minv(q) - 
          dh_{p_u} (\Id{n-k} \otimes p\tpose)\nabla_{q_u}\Minv(q) 
         \right)\simpleB} = k
         ,
    \]
    everywhere on the constraint manifold \(\Gamma\).
\end{thm}
\begin{proof}
    \textbf{TODO: prove this in a terse way}.
\end{proof}

>>>>

\subsection{Notation}
We use the following notation.
For \(T > 0\), the set of real numbers modulo \(T\) is denoted \(\Rt{T}\), with
\(\Rt{\infty} := \R\).
The gradient of the matrix-valued function 
\(A : \R^m \rightarrow \R^{n\times n}\) is the matrix of stacked partial
derivatives, 
\(\nabla_xA := (\pdiff{A}{x_1},\ldots,\pdiff{A}{x_m})\tpose \in \R^{nm \times n}\).
Given two matrices \(A \in \R^{n \times m}\) and \(B \in \R^{r \times s}\), the
Kronecker product \cite{kronprod} \(A \otimes B \in \R^{nr \times ms}\) is the
matrix
\[
    A \otimes B = \begin{bmatrix}
        A_{1,1}B & \cdots & A_{1,m} B \\
        \vdots & \ddots & \vdots \\
        A_{n,1} B & \cdots & A_{n,m} B
    \end{bmatrix} 
    .
\]

@@

\subsection{Simply Actuated Hamiltonian Systems}
Take an underactuated mechanical system modelled with generalized coordinates 
\(q = (q_1, \ldots, q_n)\) on the configuration manifold
\(\mathcal{Q} = \Rt{T_1} \times \cdot \Rt{T_n}\), where
\(T_i = 2\pi\) if \(q_i\) is an angle and \(T_i = \infty\) if \(q_i\) is a
displacement. The corresponding velocities are 
\(\dot{q} = (\dot{q}_1,\ldots,\dot{q}_n) \in \R^n\).

Suppose this system has Lagrangian
\(\mathcal{L}(q,\dot{q}) = 1/2 \dot{q}^T D(q) \dot{q} - P(q)\),
where the potential energy 
\(P : \mathcal{Q} \rightarrow \mathbb{R}\) 
is smooth, and the inertia matrix 
\(D : \mathcal{Q} \rightarrow \mathbb{R}^{n \times n}\)
is positive definite for all \(q \in \mathcal{Q}\).
Define the \textit{conjugate of momentum} 
\(p_i := \partial\mathcal{L}/\partial\dot{q}_i\) so that the vector 
\(p = (p_1,\ldots,p_n) \in \R^n\) is given by \(p = D(q)\dot{q}\).
The \textit{Hamiltonian} of the system in \((q,p)\) coordinates
is\cite{landau_mechanics}
\begin{equation}\label{eqn:hamiltonian}
    \mathcal{H}(q,p) = 1/2 p\tpose D(q) p + P(q).
\end{equation}
with dynamics
\begin{equation}\label{eqn:hamiltonian-eom-general}
    \begin{cases}
        \dot{q} = \nabla_p\mathcal{H} 
        , \\
        \dot{p} = -\nabla_q\mathcal{H} + B(q) \tau
        ,
    \end{cases}
\end{equation}
where \(\tau \in \R^k\) is a vector of generalized input forces and the input
matrix \(B : \mathcal{Q} \rightarrow \R^{n \times k}\) is full rank for all 
\(q \in \mathcal{Q}\).

@@

Because \(\tau\) is transformed by \(B(q)\), it is not in general clear how any
particular input force \(\tau_i\) will affect the dynamics of the system. 
Using the following assumptions on the input matrix, we will show there is
a canonical coordinate transformation of ~\eqref{eqn:hamiltonian} 
which makes explicit the effect of input forces.

@@

In light of Lemma \ref{lemma:B-orthogonal}, there is no loss of generality in
assuming that the input matrix is left semi-orthogonal.
Let now \(\mathbf{B} \in \R^{n\times n}\) be the following matrix:
\[
    \mathbf{B} = 
    \begin{bmatrix}
        B^\perp \\
        B\tpose \\
    \end{bmatrix}
    .
\]
Since \(B^\perp\) is a left annihilator of \(B\) and both \(B^\perp\) and
\(B\tpose\) are right semi-orthogonal, it is easy to show that \(\mathbf{B}\) is
orthogonal.

\begin{thm}\label{thm:simply-actuated}
    Take the Hamiltonian system ~\eqref{eqn:hamiltonian}.
    Under Assumptions \ref{assm:B-const} and \ref{assm:B-perp},
    the coordinate transformation
    \(\left(\tilde{q} = \mathbf{B}q, \tilde{p} = \mathbf{B}p\right)\)
    is a canonical transformation.
    The resulting dynamics are given by 
    \begin{gather}\label{eqn:simple-hamiltonian}
        \mathcal{H}(\tilde{q},\tilde{p}) = 
        \frac{1}{2} \tilde{p}\tpose \Minv(\tilde{q}) \tilde{p} + V(\tilde{q})
        , \\
       \begin{cases}
           \dot{\tilde{q}} = \Minv(\tilde{q})\tilde{p}
           , \\
           \dot{\tilde{p}} = -\frac{1}{2} (\Id{n} \otimes \tilde{p}\tpose)
           \nabla_{\tilde{q}} \Minv(\tilde{q}) \tilde{p}
           - \nabla_{\tilde{q}} V(\tilde{q}) + \simpleB \tau
            ,
        \end{cases} \nonumber
    \end{gather}
    where 
    \(\Minv(\tilde{q}) := \mathbf{B}D^{-1}(\mathbf{B}\tpose
    \tilde{q})\mathbf{B}\tpose\)
    and
    \(V(\tilde{q}) := P(\mathbf{B}\tpose \tilde{q})\).
\end{thm}
\begin{proof}
    \textbf{TODO: prove this in a terse way.}
\end{proof}
These new coordinates are called \textit{simply actuated coordinates}.
The first \((n-k)\) coordinates of \(\tilde{q}\), labelled \(q_u\), are called
the \textit{unactuated coordinates}. 
The remaining \(k\) coordinates, labelled \(q_a\), are called the
\textit{actuated coordinates}.
The corresponding \((p_u, p_a)\) are called the \textit{unactuated} and 
\textit{actuated momenta}, respectively.

Throughout the rest of this paper, we will assume the system we are studying is
in simply actuated coordinates.

@@

\begin{defn}\label{defn:vnhc}
    A \textit{virtual nonholonomic constraint} (VNHC) \textit{of order \(k\)} is a
    relation \(h(q,p) = 0\) where \(h : \mathcal{Q}\times\R^n \rightarrow \R^k\) is
    \(C^2\), \(\rank{\left[ dh_q,\, dh_p \right]} = k\) for all 
    \((q,p) \in h\inv(0)\), and there exists a feedback controller \(\tau(q,p)\)
    rendering the \textit{constraint manifold} \(\Gamma\) invariant,
    where
    \[
        \Gamma = \left\{(q,p) \mid h(q,p) = 0, dh_q \dot{q} + dh_p \dot{p} = 0\right\}
        .
    \]
\end{defn}

Definition \ref{defn:vnhc} implies that the constraint manifold \(\Gamma\)
is a \(2(n-k)\)-dimensional closed embedded submanifold of 
\(\mathcal{Q} \times \R^n\).

Define the error output \(e = h(q,p)\) 
%TODO: tie this together

\begin{thm}\label{thm:vnhc-regularity}
    A relation \(h(q,p) = 0\) for system ~\eqref{eqn:simple-hamiltonian}
    is a regular VNHC of order \(k\) if and only if \(dh_{p_a} = 0\) 
    and
    \[
        \rank{\left(dh_q \Minv(q) - 
          dh_{p_u} (\Id{n-k} \otimes p\tpose)\nabla_{q_u}\Minv(q) 
         \right)\simpleB} = k
         ,
    \]
    everywhere on the constraint manifold \(\Gamma\).
\end{thm}
\begin{proof}
    \textbf{TODO: prove this in a terse way}.
\end{proof}
